\doctitle

### 2023-04-10 msr3.tex
Add theory paper to the thesis

Defining group-specific error terms $\omega_i = Z_i u_i + \varepsilon_i$, we get a compact formulation that 
recasts~\eqref{eq:lme_setup} as a correlated noise model:
 \eq{
 \label{eq:lmm_correlated_noise_setup}
	Y_i = X_i\beta + \omega_i,\quad \omega_i \sim \NN(0, \Omega_i(\Gamma)), \quad \Omega_i(\Gamma) = Z_i\Gamma Z_i^T + \Lambda_i.
}
For brevity, we refer to $\Omega_i(\Gamma)$ as just $\Omega_i$.
%without the parentheses. In fact, $\Omega_i$ will be the only terms depending on $\Gamma$ in the majority of expressions including the mixed model's likelihood and its derivatives described below.
The reformulation \eqref{eq:lmm_correlated_noise_setup}
yields the following marginalized  negative log-likelihood function of a linear mixed-effects model~\citep{Patterson1971}:
\eq{
	\label{eq:lmm_objective}
	\LL_{ML}(\beta, \Gamma)  :=
%	 \textabove{$\Delta$}{=}
	 \sum_{i = 1}^m \half(y_i - X_i\beta)^T\Omega_i^{-1}(y_i - X_i\beta) + \half\ln{\det{\Omega_i}}.
}
Maximum likelihood estimates for $\beta$ and $\Gam$ are obtained by
solving the optimization problem
\eq{
	\label{eq:ml_lme_optimization_setup}
	\min_{\beta, \Gamma} & \ \LL_{ML}(\beta, \Gamma)  \quad \mbox{s.t.} \quad \ \Gamma \in \bS_{+}^{q}.
}
At this point, we bring in three basic definitions from variational analysis~\cite{rockafellar2009variational}.

@@

Table \ref{table:proxes} provides closed form expressions for the proximal operators 
of commonly used regularizers. 
%All regularizers in this list are separable except $R(x) = \delta_{\|x\|_0 \leq k}$, for which the solution also has an analytically closed form.
%Again when $R$ is separable with the $r_i$ coming from Table \ref{table:proxes}, 
%the evaluation of 
For all of these cases,  the following theorem gives closed form expressions for
$\prox_{\alpha R + \delta_{\CC}}(z)$.

\begin{theorem}[$\prox$ for bounded $\gamma$]     \label{thm:prox_of_positive_quadrant}
We consider modified regularizers $r(\gamma)$ from the Table \ref{table:proxes} 
% and let $v = (\beta, \gamma)$. % \in \CC \subset R^{p + q}$.
that include an additional constraint on $\gamma$ of the form 
\[
0 \leq \gamma \leq \bar\gamma,
\]
for $\bar\gamma\in[0,+\infty]$.
%which which the positive orthant as a special case (i.e. $\bar\gamma = \infty$). 
We have the following results. 
%    \begin{enumerate}
%    \item  For $R(x)$ given by LASSO, A-LASSO, CAD, and SCAD, we have  for all $i$ that 
    %\eq{
    %\prox_{\alpha R + \delta_{\CC}}(v) = (\prox_{\alpha r}(v_1), \dots, \prox_{\alpha r}(v_p), \prox_{\alpha r + \delta_{\R_+}}(v_{p+1}), \dots, \prox_{\alpha r + \delta_{\R_+}}(v_{p+q}))
    %}
    %where 
 %   \eq{
 %   \prox_{\alpha r + \delta_{\R_+}}(v_{i}) = \begin{cases} \prox_{\alpha r}(v_i), & v_i \geq 0 \\ 0, & \text{ otherwise} \end{cases} 
%    \item For $R(x) = \delta_{\|x\|_0 \leq k}$, the $\prox_{\alpha R + \delta_{\CC}}(v)$ can be evaluated by taking $k$ largest non-negative coordinates of $v$, and setting the rest to $0$.
 %   \end{enumerate}
%    Next, suppose $\LL = \LL_{REML}$. 
    \begin{enumerate}
    \item  For SCAD, we have for all $i$ that  
   % \eq{
%    \prox_{\alpha R + \delta_{\CC}}(v) = (\prox_{\alpha r}(v_1), \dots, \prox_{\alpha r}(v_p), \prox_{\alpha r + \delta_{\R_+} + \delta_{\R_{\leq \bar\gamma}}}(v_{p+1}), \dots, \prox_{\alpha r + \delta_{\R_+} + \delta_{\R_{\leq \bar\gamma}}}(v_{p+q}))
 %   }
  %  where 
%    \eq{
\[
    \prox_{(\alpha r + \del_{[0,\bgam]})} %\delta_{\R_+} + \delta_{\R_{\leq \bar\gamma}}}
    (\gamma_{i}) = \begin{cases} 
        \prox_{\alpha r}(\gamma_i), & 0 \leq \gamma_i < \bar\gamma  \\
        \bar\gamma, & \gamma_i \geq \bar\gamma \\
        0, & \text{ otherwise} \end{cases}
\]
%    }
    \item  For LASSO, A-LASSO we have for all $i$ that 
 %   \eq{
  %  \prox_{\alpha R + \delta_{\CC}}(v) = (\prox_{\alpha r}(v_1), \dots, \prox_{\alpha r}(v_p), \prox_{\alpha r + \delta_{\R_+} + \delta_{\R_{\leq \bar\gamma}}}(v_{p+1}), \dots, \prox_{\alpha r + \delta_{\R_+} + \delta_{\R_{\leq \bar\gamma}}}(v_{p+q}))
  %  }
  %  where 
%    \eq{
\[
\prox_{(\alpha r + \del_{[0,\bgam]})}(\gamma_i) 
%    \prox_{\alpha r + \delta_{\R_+} + \delta_{\R_{\leq \bar\gamma}}}(v_{i}) 
    = \begin{cases} 
        \prox_{\alpha r}(\gamma_i), & 0 \leq \gamma_i < \bar\gamma + \alpha \\
        \bar\gamma, & \gamma_i \geq \bar\gamma + \alpha \\
        0, & \text{ otherwise} \end{cases}
\]
%    }
    \item For $R(\cdot) = \delta_{\lev{\norm{\cdot}_0}{k}}$  %\delta_{\|x\|_0 \leq k}$, 
    the $\prox_{\alpha R + \delta_{\CC}}(\gamma)$ can be evaluated by taking $k$ largest coordinates of $\gamma$ such that $0 \leq \gamma_i \leq \bar\gamma$, and setting the 
    remainder to $0$.
    \end{enumerate}
\end{theorem}
The proof of the Theorem \ref{thm:prox_of_positive_quadrant} is provided in Appendix \ref{ch:proxes_appendix}. 
%\textcolor{red}{Aleksei, please check proof of appendix is all in terms of $\gamma$, rather than $v = (\beta, \gamma)$.}
The  PGD algorithm is detailed in Algorithm~\ref{alg:pgd_for_lme}.
{The algorithm's step-size $\alpha$ depends on the Lipschitz constant; an upper-bound is given in Appendix \ref{sec:lipschitz_constant}. In practice, $\alpha$ is computed using a line-search, since the available estimate for $L$ is very conservative.}

@@

To understand the algorithm and logic behind the objective~\eqref{eq:lme_loss_sr3}, we define a value function $u_\eta(w)$ and the solution set $S_\eta(w)$:
\eq{
\label{eq:value_function_definition}
		u_\eta(w) & = \min_x \ \LL(x) + \delta_{\CC}(x) + \kappa_{\eta}(x-w)\\
		S_\eta(w) & = \argmin_x \LL(x) + \delta_{\CC}(x) + \kappa_{\eta}(x-w).
}
Substituting (\ref{eq:value_function_definition}) into (\ref{eq:lme_loss_sr3}) the problem transforms into the problem of optimizing a regularized value function:
\eq{
	\label{eq:value_function_optimization}
	\min_{w} u_\eta(w) + R(w)
}
%Thus at a conceptual level, 
Here we have transformed the original regularized likelihood~\eqref{eq:lme_loss_original_in_x}  %has been transformed 
through relaxation and partial 
minimization to obtain an equivalent problem~\eqref{eq:value_function_optimization} 
for $w$ with the same regularizer. The value function $u_\eta$ encapsulates 
global variational information on the function $\LL(x) + \delta_{\CC}(x)$
relative to $w$.
%Problem~\eqref{eq:value_function_optimization} 
%attempts to find sparse $w$ that are small with respect to $u_\eta$.

@@

\section{Theoretical Analysis}
\ask{Jim}{What would be the best way to introduce theoretical results to my thesis?}

>>>>

Defining group-specific error terms $\omega_i = Z_i u_i + \varepsilon_i$, we get a compact formulation that 
recasts~\eqref{eq:lme_setup} as a correlated noise model:
 \eq{
 \label{eq:lmm_correlated_noise_setup}
	Y_i = X_i\beta + \omega_i,\quad \omega_i \sim \NN(0, \Omega_i(\Gamma)), \quad \Omega_i(\Gamma) = Z_i\Gamma Z_i^T + \Lambda_i.
}
For brevity, we refer to $\Omega_i(\Gamma)$ as just $\Omega_i$.
%without the parentheses. In fact, $\Omega_i$ will be the only terms depending on $\Gamma$ in the majority of expressions including the mixed model's likelihood and its derivatives described below.
The reformulation \eqref{eq:lmm_correlated_noise_setup}
yields the following marginalized  negative log-likelihood function of a linear mixed-effects model~\citep{Patterson1971}:
\eq{
	\label{eq:lmm_objective}
	\LL_{ML}(\beta, \Gamma)  :=
%	 \textabove{$\Delta$}{=}
	 \sum_{i = 1}^m \half(y_i - X_i\beta)^T\Omega_i^{-1}(y_i - X_i\beta) + \half\ln{\det{\Omega_i}}.
}
Maximum likelihood estimates for $\beta$ and $\Gam$ are obtained by
solving the optimization problem
\eq{
	\label{eq:ml_lme_optimization_setup}
	\min_{\beta, \Gamma} & \ \LL_{ML}(\beta, \Gamma)  \quad \mbox{s.t.} \quad \ \Gamma \in \bS_{+}^{q}.
}

\begin{theorem}[Existence of a Minimizer, Theorem 1 from \cite{aravkin2022jimtheory}]\label{thm:basic existence}
Let the assumptions in the statement of problem \eqref{eq:ml_lme_optimization_setup} hold. Then optimal solutions to
\eqref{eq:ml_lme_optimization_setup} exist.
\end{theorem}

At this point, we bring in three basic definitions from variational analysis~\cite{rockafellar2009variational}.

@@

Under the assumption that $R$ is level-compact \cite{aravkin2022jimtheory} extended Theorem \ref{thm:basic existence} to the regularized case as follows:

\begin{theorem}[Existence of a Minimizer for \ref{eq:lme_loss_original_in_x}, \cite{aravkin2022jimtheory}, Theorem 2]

\label{thm:basic existence2}
Let the assumptions in the statement of problem \eqref{eq:lme_diagonal_setup} hold.
%define $\LL(\beta,\gam):=\LL_{ML}(\beta, \Diag{\gamma})$
Suppose $\map{\hR}{\R^p\times\R^q_+}{\R\cup\{+\infty\}}$ 
is lsc and level compact (i.e., $\epi{R}:=\lset{(\bg,\nu)}{R\bg\le\nu}$ is closed
and $\lset{\bg}{R\bg\le\nu}$ is bounded for all $\nu\in\R$).
Then $\LL+\hR$ is level compact and solutions to
the following optimization problem exist:
\eq{
\label{eq:extended loss}
\min_{\beta\in\R^p, \gamma\in\R^q_+}  \LL(\beta, \gamma) + 
\hR(\beta, \gamma) .
}
\end{theorem}

In practice, it is often advisable to include a constraint of the form 
$\gamma\le\gammax$ for $\gam\in\R^q_{++}$ chosen
sufficiently large since an excessively large variance usually indicates 
that the model is poorly posed and needs review. 
Such a constraint is also numerically expedient since it prevents $\gam$ from diverging. Table \ref{table:proxes} provides closed form expressions for the proximal operators 
of commonly used regularizers. 
%All regularizers in this list are separable except $R(x) = \delta_{\|x\|_0 \leq k}$, for which the solution also has an analytically closed form.
%Again when $R$ is separable with the $r_i$ coming from Table \ref{table:proxes}, 
%the evaluation of 
For all of these cases,  the following theorem gives closed form expressions for
$\prox_{\alpha R + \delta_{\CC}}(z)$.

\begin{theorem}[Proximal operator for bounded $\gamma$]     \label{thm:prox_of_positive_quadrant}
We consider modified regularizers $r(\gamma)$ from the Table \ref{table:proxes} 
% and let $v = (\beta, \gamma)$. % \in \CC \subset R^{p + q}$.
that include an additional constraint on $\gamma$ of the form 
\[
0 \leq \gamma \leq \bar\gamma,
\]
for $\bar\gamma\in[0,+\infty]$.
%which which the positive orthant as a special case (i.e. $\bar\gamma = \infty$). 
We have the following results. 
%    \begin{enumerate}
%    \item  For $R(x)$ given by LASSO, A-LASSO, CAD, and SCAD, we have  for all $i$ that 
    %\eq{
    %\prox_{\alpha R + \delta_{\CC}}(v) = (\prox_{\alpha r}(v_1), \dots, \prox_{\alpha r}(v_p), \prox_{\alpha r + \delta_{\R_+}}(v_{p+1}), \dots, \prox_{\alpha r + \delta_{\R_+}}(v_{p+q}))
    %}
    %where 
 %   \eq{
 %   \prox_{\alpha r + \delta_{\R_+}}(v_{i}) = \begin{cases} \prox_{\alpha r}(v_i), & v_i \geq 0 \\ 0, & \text{ otherwise} \end{cases} 
%    \item For $R(x) = \delta_{\|x\|_0 \leq k}$, the $\prox_{\alpha R + \delta_{\CC}}(v)$ can be evaluated by taking $k$ largest non-negative coordinates of $v$, and setting the rest to $0$.
 %   \end{enumerate}
%    Next, suppose $\LL = \LL_{REML}$. 
    \begin{enumerate}
    \item  For SCAD, we have for all $i$ that  
   % \eq{
%    \prox_{\alpha R + \delta_{\CC}}(v) = (\prox_{\alpha r}(v_1), \dots, \prox_{\alpha r}(v_p), \prox_{\alpha r + \delta_{\R_+} + \delta_{\R_{\leq \bar\gamma}}}(v_{p+1}), \dots, \prox_{\alpha r + \delta_{\R_+} + \delta_{\R_{\leq \bar\gamma}}}(v_{p+q}))
 %   }
  %  where 
%    \eq{
\[
    \prox_{(\alpha r + \del_{[0,\bgam]})} %\delta_{\R_+} + \delta_{\R_{\leq \bar\gamma}}}
    (\gamma_{i}) = \begin{cases} 
        \prox_{\alpha r}(\gamma_i), & 0 \leq \gamma_i < \bar\gamma  \\
        \bar\gamma, & \gamma_i \geq \bar\gamma \\
        0, & \text{ otherwise} \end{cases}
\]
%    }
    \item  For LASSO, A-LASSO we have for all $i$ that 
 %   \eq{
  %  \prox_{\alpha R + \delta_{\CC}}(v) = (\prox_{\alpha r}(v_1), \dots, \prox_{\alpha r}(v_p), \prox_{\alpha r + \delta_{\R_+} + \delta_{\R_{\leq \bar\gamma}}}(v_{p+1}), \dots, \prox_{\alpha r + \delta_{\R_+} + \delta_{\R_{\leq \bar\gamma}}}(v_{p+q}))
  %  }
  %  where 
%    \eq{
\[
\prox_{(\alpha r + \del_{[0,\bgam]})}(\gamma_i) 
%    \prox_{\alpha r + \delta_{\R_+} + \delta_{\R_{\leq \bar\gamma}}}(v_{i}) 
    = \begin{cases} 
        \prox_{\alpha r}(\gamma_i), & 0 \leq \gamma_i < \bar\gamma + \alpha \\
        \bar\gamma, & \gamma_i \geq \bar\gamma + \alpha \\
        0, & \text{ otherwise} \end{cases}
\]
%    }
    \item For $R(\cdot) = \delta_{\lev{\norm{\cdot}_0}{k}}$  %\delta_{\|x\|_0 \leq k}$, 
    the $\prox_{\alpha R + \delta_{\CC}}(\gamma)$ can be evaluated by taking $k$ largest coordinates of $\gamma$ such that $0 \leq \gamma_i \leq \bar\gamma$, and setting the 
    remainder to $0$.
    \end{enumerate}
\end{theorem}
The proof of the Theorem \ref{thm:prox_of_positive_quadrant} is provided in Appendix \ref{appendix:proxes}. 
%\textcolor{red}{Aleksei, please check proof of appendix is all in terms of $\gamma$, rather than $v = (\beta, \gamma)$.}
The  PGD algorithm is detailed in Algorithm~\ref{alg:pgd_for_lme}.
{The algorithm's step-size $\alpha$ depends on the Lipschitz constant; an upper-bound is given in Appendix \ref{appendix:lipschitz_constant}. In practice, $\alpha$ is computed using a line-search, since the available estimate for $L$ is very conservative.}

@@

To understand the algorithm and logic behind the objective~\eqref{eq:lme_loss_sr3}, we define a value function $u_\eta(w)$ and the solution set $S_\eta(w)$:
\eq{
		u_\eta(w) & = \min_x \ \LL(x) + \delta_{\CC}(x) + \kappa_{\eta}(x-w)\\
		S_\eta(w) & = \argmin_x \LL(x) + \delta_{\CC}(x) + \kappa_{\eta}(x-w).
}
Substituting the above into (\ref{eq:lme_loss_sr3}) the problem transforms into the problem of optimizing a regularized value function:
\eq{
	\label{eq:value_function_optimization}
	\min_{w} u_\eta(w) + R(w)
}
%Thus at a conceptual level, 
Here we have transformed the original regularized likelihood~\eqref{eq:lme_loss_original_in_x}  %has been transformed 
through relaxation and partial 
minimization to obtain an equivalent problem~\eqref{eq:value_function_optimization} 
for $w$ with the same regularizer. The value function $u_\eta$ encapsulates 
global variational information on the function $\LL(x) + \delta_{\CC}(x)$
relative to $w$.
%Problem~\eqref{eq:value_function_optimization} 
%attempts to find sparse $w$ that are small with respect to $u_\eta$.

@@

\section{Theoretical Analysis}
\input{msr3-theory}
